---
title: "pLoT model vs. SLM"
author: "Anneloes de Moel"
date: "2024-04-17"
output: html_document
---

## Loading the necessary packages

```{r}
library(tidyverse)
library(dplyr)
library(readr)
library(stringr)
library(lme4)
library(ggplot2)
library(flextable)
library(pROC)
library(caret)
```

## Loading the dataframe

```{r}
setwd("/Users/anneloesdemoel/Documents/UvA/Scriptie")

study1B <- read_delim("Study1B_results.csv", 
    delim = "\t", escape_double = FALSE, 
    trim_ws = TRUE)

study1B <- read_table("Study1B_results.csv")
```

## Re-creating the Streak Length Model

```{r final model}

# Convert 'type' variable to a factor
study1B$type <- factor(study1B$type)

# Extract the last number in each sequence
study1B$last_number <- as.numeric(substr(study1B$sequence, nchar(study1B$sequence), nchar(study1B$sequence)))

# Add total count occurrences
study1B$last_number_count <- str_count(study1B$sequence, substr(study1B$sequence, nchar(study1B$sequence), nchar(study1B$sequence)))

# Recreate the Streak Length Model
SLM_logit_model <- glmer(prediction_recode ~ terminal_streak_length + last_number_count + generator + (1 | participant_id), data = study1B, family = binomial)

# To determine if the predictors have a significant effect on the dependent variable
anova(SLM_logit_model) # determines that last_number_count has a higher effect than streak_length and generator has an even lower effect but together a better effect

# Summary of the Streak Length Model
summary(SLM_logit_model)

# Predict probabilities for each participant
study1B$predicted_probabilities <- predict(SLM_logit_model, type = "response")

# Changing the prediction probabilities to the probability of printing a '0'

# Function to transform predicted probabilities to probability of predicting '0' as the 9th variable
transform_prob_to_zero <- function(last_number, predicted_probabilities) {
  if (last_number == 0) {
    return(predicted_probabilities)
  } else {
    return(1 - predicted_probabilities)
  }
}

# Apply the function to each row in the dataset
study1B$prob_of_zero <- mapply(transform_prob_to_zero, study1B$last_number, study1B$predicted_probabilities)

# View the data with transformed probabilities
print(study1B)

summary(SLM_logit_model)

```

## Plotting the predicted probabilities from the Streak Length Model

```{r plotting the predicted probabilities}

# Plot the predicted probabilities against last_number_count
ggplot(study1B, aes(x = last_number_count, y = predicted_probabilities)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  labs(x = "Streak", y = "Probability of Repetition") +
  theme_minimal() +
   theme(
    axis.title.x = element_text(size = 17),  
    axis.title.y = element_text(size = 17),  
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)  
  )

# Plot the predicted probabilities against terminal_streak_length
ggplot(study1B, aes(x = terminal_streak_length, y = predicted_probabilities)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  labs(x = "Streak", y = "Probability of Repetition") +
  theme_minimal() +
   theme(
    axis.title.x = element_text(size = 17),  
    axis.title.y = element_text(size = 17),  
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)  
  )
```

## Making a sample table

```{r making a sample table}

sample <- study1B[sample(nrow(study1B), 10), ] %>%
  print(width = Inf)

# Select relevant columns
sample_selected <- sample %>%
  select(participant_id, generator, sequence, terminal_streak_length, prediction_recode, last_number_count, predicted_probabilities)

# Create the flextable
flextable_table <- flextable(sample_selected) %>%
  set_header_labels(
    participant_id = "Participant ID",
    generator = "Generator",
    sequence = "Sequence",
    terminal_streak_length = "Terminal Streak Length",
    prediction_recode = "Prediction Recode",
    last_number_count = "Last Number Count",
    predicted_probabilities = "Predicted Probabilities"
  ) %>%
  bold(part = "header") %>%
  align(align = "center", part = "all")

# Print the flextable
flextable_table
```

## Loading the pLOT model outputs

```{r}
# Load the dataset containing 'sequences' and 'predicted' (the probabilities)
better_probs <- read.csv("/Users/anneloesdemoel/Documents/UvA/Scriptie/Python/better_probs.csv")

# Fill the shorter sequences with 0s to get length 8
better_probs$sequence <- str_pad(better_probs$sequence, width = 8, side = "left", pad = "0")

# Join the datasets by sequence
comparison <- inner_join(
  study1B,
  better_probs,
  by = "sequence"
)
```

## Transforming pLOT predictions into probability of repetition

```{r}
# transform the pLoT probabilities to probability of repetition
transform_prob_to_repeat <- function(last_number, prediction) {
  if (last_number == 0) {
    return(prediction)
  } else {
    return(1 - prediction)
  }
}

# Apply transformation to all results
comparison$pLOT_repeat <- mapply(transform_prob_to_repeat, comparison$last_number, comparison$prediction)

print(comparison)

# Apply pLoT model results to the initial Streak Length Model as predictor variable
pLOT_logit_model <- glmer(prediction_recode ~ pLOT_repeat + terminal_streak_length + last_number_count + generator + (1 | participant_id), 
                     data = comparison, 
                     family = binomial)

comparison$pLOT_prediction <- predict(pLOT_logit_model, type = "response")

summary(pLOT_logit_model) # determine whether the addition of the results create a more significant model for predicting repetition
```

## Plotting the pLOT recoded

```{r}
comparison$prediction_recode <- as.factor(comparison$prediction_recode)

ggplot(comparison, aes(x = prediction_recode, y = pLOT_prediction)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.1) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  labs(x = "Recode", y = "Predicted", title = "pLOT repetition") +
  theme_minimal() +
   theme(
    axis.title.x = element_text(size = 14),  
    axis.title.y = element_text(size = 14),  
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)  
  )
```

## Ploting the predicted probability of repetition against Rao & Hastie recoded probabilities

```{r}
comparison$prediction_recode <- as.factor(comparison$prediction_recode)

ggplot(comparison, aes(x = prediction_recode, y = predicted_probabilities)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.1, size = 1, alpha = 0.1) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  labs(x = "Recode", y = "Predicted", title = "SLM repetition") +
  theme_minimal() +
   theme(
    axis.title.x = element_text(size = 14),  
    axis.title.y = element_text(size = 14),  
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)  
  )
```

```{r}
library(ggplot2)
library(dplyr)

# Create a combined data frame of the predicted probabilities
combined_data <- comparison %>%
  gather(key = "type", value = "prediction", pLOT_prediction, predicted_probabilities) %>%
  mutate(type = ifelse(type == "pLOT_prediction", "pLOT repetition", "SLM repetition"))

# Ensure prediction_recode is a factor
combined_data$prediction_recode <- as.factor(combined_data$prediction_recode)

# Plot the comparison
ggplot(combined_data, aes(x = prediction_recode, y = prediction, fill = type)) +
  geom_violin(trim = FALSE, position = position_dodge(width = 0.9)) +
  geom_jitter(aes(color = type), width = 0.1, size = 1, alpha = 0.05) +
  geom_smooth(aes(group = type, color = type), method = "glm", method.args = list(family = "binomial"), se = FALSE, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("pLOT repetition" = "#A9A9A9", "SLM repetition" = "#D3D3D3")) +
  scale_color_manual(values = c("pLOT repetition" = "#696969", "SLM repetition" = "#B0B0B0")) +
  labs(x = "Recode", y = "Predicted", title = "Split Violin Plot of pLoT and SLM Repetitions") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 14),  
    axis.title.y = element_text(size = 14),  
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)  
  )

ggplot(combined_data, aes(x = as.factor(prediction_recode), y = prediction, fill = type)) +
  geom_violin(trim = FALSE, position = position_dodge(width = 0.9)) +
  geom_jitter(aes(color = type), size = 1, alpha = 0.2, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9)) +
  geom_smooth(aes(group = type, color = type), method = "glm", method.args = list(family = "binomial"), se = FALSE, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("pLOT repetition" = "#A9A9A9", "SLM repetition" = "#D3D3D3")) +
  scale_color_manual(values = c("pLOT repetition" = "royalblue3", "SLM repetition" = "springgreen4")) + 
  labs(x = "Recode", y = "Predicted", title = "Split Violin Plot of pLoT and SLM Repetitions") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 14),  
    axis.title.y = element_text(size = 14),  
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )
```

# Comparing the models with a nested model comparison using ANOVA

```{r}
# First reprint the SLM and pLoT model
SLM_logit_model <- glmer(prediction_recode ~ terminal_streak_length + last_number_count + generator + (1 | participant_id), data = comparison, family = binomial)

pLOT_logit_model <- glmer(prediction_recode ~ pLOT_repeat + terminal_streak_length + last_number_count + generator + (1 | participant_id), data = comparison, family = binomial)

# Run the anova code to compare the models through Chi squared test
anova(pLOT_logit_model, SLM_logit_model, test = "Chisq") # including the pLOT results provides a much better fitting model
```

## Comparing the model outputs for predicting repetition

```{r confusion matrix for probability of repetition}

# Binarize probabilities with a threshold
threshold <- 0.5 # every value above 0.5 was classified as 'change' and every value below was classified as 'repeat'
SLM_repeat_prediction <- ifelse(comparison$predicted_probabilities > threshold, 1, 0)
pLOT_repeat_prediction <- ifelse(comparison$pLOT_prediction > threshold, 1, 0)

# Confusion Matrix for Streak Length Model
accuracy_SLM <- confusionMatrix(factor(SLM_repeat_prediction), factor(comparison$prediction_recode))
print(accuracy_SLM)

# Confusion Matric for Probabilistic Language of Thought Model
accuracy_pLOT <- confusionMatrix(factor(pLOT_repeat_prediction), factor(comparison$prediction_recode))
print(accuracy_pLOT) 
```

## Printing confusion matrices

```{r}

# Create a dataframe from the confusion matrix for LOT
conf_pLOT <- as.data.frame(accuracy_pLOT$table)

# Create a dataframe from the confusion matrix for SLM
conf_SLM <- as.data.frame(accuracy_SLM$table)

# Combine the confusion matrices into a single dataframe
conf_combined <- bind_rows(
  mutate(conf_SLM, Model = "Streak Length Model"),
  mutate(conf_pLOT, Model = " Probabilistic Language of Thought Model")
)

# Plot confusion matrices using ggplot and facets
ggplot(data = conf_combined, aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "white") +
  scale_fill_gradient(low = "white", high = "skyblue3") +
  geom_text(aes(label = Freq), vjust = 1) +
  labs(title = "Model Accuracy Comparison",
       x = "Actual",
       y = "Predicted",
       fill = "Frequency") +
  facet_wrap(~Model, ncol = 2) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 17),  
    axis.title.y = element_text(size = 17),  
    plot.title = element_text(size = 17, face = "bold", hjust = 0.5)  
  )

```
